Tropical mathematics and the Lambda-calculus I

### 2023-08-02 CSL24/app_section4.tex
Commit

@@

@@

Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative, we immediately have:

\begin{theorem}[\autoref{thmTLSlocLip}]
All tps $f:\Lawv^X\to\Lawv$ are locally Lipschitz on $(0,\infty)^X$.
Moreover, the Lipschitz constant of $f$ on $\overline{B_{\delta}(x)}$ can be chosen to be $\frac{1}{\delta}\max_{\overline{B_{3\delta}(x)}} f$.
\end{theorem}

>>>>

\subsection{Proof of \autoref{prop:nondecr+conc}}

Remember that a function $f:Q^{X}\to Q^{Y}$ is \emph{concave} if for all $\alpha\in [0,1]$, $ x ,  y  \in Q^{X}$ and $b\in Y$ 
\[
f(\alpha\cdot  x +(1-\alpha)\cdot  y  )_{b} \geq \alpha f( x )_{b} + (1-\alpha)f( y  )_{b}
\]

We now prove \autoref{prop:nondecr+conc}, i.e.\ that $f: Q^{X}\to Q^{Y}$ are non-decreasing and concave.

The fact that $f$ is non-decreasing is clear, since the multiplicities of the multisets and all coordinates of the points are non-negative.
Let us show the concavity.
Let us first show that all functions of the form $f( x )_{b}= \mu  x + c$ are concave:
we have $f(\alpha x + (1-\alpha) y  )_{b}= \mu(\alpha x )+(1-\alpha) y  )+c=
 \mu(\alpha x )+(1-\alpha) y  )+\alpha c+(1-\alpha)c=
 \alpha(\mu  x  + c)+(1-\alpha)(\mu  y  +c)=\alpha f( x )_{b}+(1-\alpha) f( x )_{b}$.
To conclude, let us show that if $(f_{i})_{i\in I}$ is a family of concave functions from $Q^{X}$ to $Q^{Y}$, the function $f=\inf_{i\in I}f_{i}$ is also concave: we have
$f(\alpha x  +(1-\alpha) y  )_{b}=
\inf_{i\in I}f_{i}(\alpha x +(1-\alpha) y  )_{b} \geq 
\inf_{i\in I}\alpha f_{i}( x )_{b}+(1-\alpha)f_{i}( y  )_{b}
\geq 
\inf_{i\in I}\alpha f_{i}( x )_{b} + \inf_{j\in I}(1-\alpha)f_{j}( y  )_{b}
=
\alpha \cdot (\inf_{i\in I}f_{i}( x )_{b})+ (1-\alpha)\cdot( \inf_{j\in I}f_{j}( y  )_{b})=
\alpha  f( x )_{b}+(1-\alpha)f( y  )_{b}$, where we used the fact that given families $a_{i},b_{i}$ of reals,
$\inf_{i}a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{j}b_{j}$.
This follows from the fact that for all $i\in I$, $a_{i}+b_{i}\geq \inf_{i}a_{i}+\inf_{i}b_{i}$.

\subsection{Proof of \autoref{thm:ScottCont}}

The part of \autoref{thm:ScottCont} about tPs immedley follows from the first part of the same theorem.
Let us quickly recall the basic definitions about cones that we need in order to prove it.

\begin{definition}
 An \emph{$\overline{\R}_{\geq 0}$-cone} is a commutative $\overline{\R}_{\geq 0}$-semimodule with cancellative addition (i.e.\ $x+y=x+y' \Rightarrow y=y'$).
\end{definition}

In \cite{Selinger2004} cones are required to also have ``strict addition'', meaning that $x+y=0 \Rightarrow x=y=0$.
We do not add this requirement since it will automatic hold when considering normed cones.

\begin{remark}
 The addition of a cone $P$ (which forms a commutative monoid) turns $P$ into a poset by setting:
 $
  x \leq y \textit{ iff } y=x+z \textit{, for some }z\in P.
 $
 This is called the \emph{cone-order} on $P$
 By the cancellative property, when such $z$ exists it is unique, and we denote it by $y-x$.
\end{remark}

\begin{definition}
 A \emph{normed $\overline{\R}_{\geq 0}$-cone} $P$ is the data of a $\overline{\R}_{\geq 0}$-cone together with a $\leq$-monotone\footnote{I.e.: $x\leq y \Rightarrow \norm{x}\leq \norm y$. Remark that requiring this property (for all $x,y$) is equivalent to requiring that $\norm{x}\leq \norm{x+y}$ for all $x,y$.} norm on it, where a \emph{norm} on $P$ is a map $\norm{.}:P\to \overline{\R}$ satisfying the usual axioms of norms:
 $\norm x \geq 0$, $\norm x = 0 \Rightarrow x=0$, $\norm{rx}=r\norm x$ and $\norm{x+y}\leq \norm x + \norm y$.
\end{definition}

In, e.g.~\cite{EhrPagTas2018}, a normed $\overline{\R}_{\geq 0}$-cone is simply called a cone.

Remark that in a normed $\overline{\R}_{\geq 0}$-cone, by monotonicity of the norm, we have: $\norm{x+y}=0 \Rightarrow x=y=0$.
Therefore, as already mentioned, in a normed cone we have: 
$x+y=0 \Rightarrow \norm{x+y}=0 \Rightarrow x=y=0$, that is, addition is strict.

\begin{example}
 $\overline{\R}_{\geq 0}^X$ is a normed cone with the norm $\supnorm{x}:=\sup\limits_{a\in X} x_a\in \overline{\R}_{\geq 0}$.
\end{example}

\begin{remark}
 The cone-order on $\overline{\R}_{\geq 0}^X$ is the pointwise usual order on $\overline{\R}_{\geq 0}$.
% Remark also that tLs have no reason to be linear nor sublinear.
\end{remark}

A \emph{directed net} in a poset $P$ with indices in a set $I$ is a function $s:I\to P$, denoted by $(s_i)_{i\in I}$, s.t.\ its image is directed.
We say that a directed net in $P$ \emph{admits a sup} iff its image admits a sup in $P$.
We say that a directed net $s$ in a normed cone is \emph{bounded} iff the set $\set{\norm{s_i}\,\mid i\in I}$ is bounded in $\R_{\geq 0}$.

Remember the definition of Scott-continuity:

\begin{definition}
 A function $f:P\to P'$ between posets is \emph{Scott-continuous} iff for all directed net $(s_i)_i$ in $P$ admitting a sup, we have $\exists \bigvee\limits_i f(s_i) = f(\bigvee\limits_i s_i)$ in $P'$. 
\end{definition}

The fundamental result in order to prove Theorem~\ref{thm:ScottCont} is the following, taken from \cite{Selinger2004}.

\begin{proposition}\label{prop:infsup}
 Let $P$ be a normed $\overline{\R}_{\geq 0}$-cone s.t.\ every bounded directed net in $P$ admits a sup.
 Let $(v_i)_{i\in I}$ be a directed net in $P$ with an upper bound $v\in P$.
 Then $\exists\bigvee\limits_{i\in I} v_i \in P$ and, if $\inf\limits_{i\in I} \norm{v-v_i} =0$, one has: $\bigvee\limits_{i\in I} v_i = v$.
\end{proposition}
\begin{proof}
 Remark that $v-v_i$ exists in $P$ by hypothesis and so does $\bigvee\limits_{i\in I} v_i$. %, thanks to the monotonicity of the norm.
 Now, since $v\geq v_i$ for all $i$, we have that $v\geq \bigvee\limits_{i\in I} v_i$, and so $v-\bigvee\limits_{i\in I} v_i$ exists in $P$.
 Fix $i\in I$.
 Since $v_i\leq \bigvee\limits_{i\in I} v_i$, then $v-\bigvee\limits_{i\in I} v_i\leq v-v_i$ and, by monotonicity of the norm, $\norm{v-\bigvee\limits_{i\in I} v_i}\leq \norm{v-v_i}$.
 Since this holds for all $i\in I$, we have:
 $0\leq \norm{v-\bigvee\limits_{i\in I} v_i}\leq \inf\limits_{i\in I} \norm{v-v_i}=0$, where the last equality holds by hypothesis.
 Thus $\norm{v-\bigvee\limits_{i\in I} v_i}=0$, i.e.\ $v=\bigvee\limits_{i\in I} v_i$.
\end{proof}

We finally obtain the desired:

\begin{theorem}[Theorem~\ref{thm:ScottCont}]
  All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
\end{theorem}
\begin{proof}
 Let $(x_i)_i$ a directed net in $(0,\infty)^X$ s.t.\ $\bigvee\limits_i x^i$ exists in $(0,\infty)^X$.
 Then $\inf\limits_i \supnorm{\bigvee\limits_i x^i - x^i} =0$, where $\bigvee\limits_i x^i - x^i$ exists because $\bigvee\limits_i x^i \geq x^i$ for all $i$.
 Since $f$ is $\supnorm{.}$-continuous on $(0,\infty)^X$, then $\inf\limits_i \supnorm{f(\bigvee\limits_i x^i) - f(x^i)} =0$, where $f(\bigvee\limits_i x^i) - f(x^i)$ exists because $f(\bigvee\limits_i x^i) \geq f(x^i)$ for all $i$ being $f$ monotone.
 We can therefore apply Proposition \ref{prop:infsup} to the directed net $(f(x^i))_i$ in $(0,\infty)$, obtaining that $\bigvee\limits_i f(x^i)$ exists in $(0,\infty)$ and it coincides with $f(\bigvee\limits_i x^i)$.
\end{proof}

\subsection{Proof of \autoref{thmTLSlocLip}}

@@

The main ingredient of the proof, that we mention in the proof sketch of \autoref{thmTLSlocLip}, is the following:

@@

Therefore, since $(0,\infty)^X$ is open and convex in $(\BB R^X,\norm\cdot)$ and all tps are non-negative, we immediately have \autoref{thmTLSlocLip}.

### 2023-08-02 CSL24/section4.tex
Commit

We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} (see~\cite{Selinger2004}, or the appendix, for such notions).
Its cone structure induces an order on it, called the \emph{cone order}:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
Such order actually coincides with the pointwise order on $\Lawv^X$.
This makes it a Scott-continuous dcpo.
Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
In analogy with \cite[Proposition 17]{DanEhrh2011}, we have:

\begin{theorem}\label{thm:ScottCont}
 All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
 In particular, all tps $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
\end{theorem}
%\begin{proof}
% From \autoref{thm:cont} plus the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
%\end{proof}

@@

\begin{proposition}\label{prop:troplinear}
\begin{enumerate}
\item If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:X\times Y\to \Lawv$ (i.e.~it is linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).  
\item If  $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
\end{enumerate}
\end{proposition}
\begin{proof}[Proof sketch]
1). Using the fact that $f( x)_{b}= \inf_{a\in X}\{\matr f_{a,b}+ x_{a}\}$,
the problem reduces to checking that $|(\matr f_{a,b}- x_{a})- (\matr f_{a,b}- y_{a})| = |x_{a}- y_{a}|\leq \|  x- y\|_{\infty}$ (see the Appendix).

>>>>

We conclude this subsection by noticing that, as a set, $\Lawv=[0,\infty]$, and  $[0,\infty]^X$ with the usual $+$ and $\cdot$ is a $\BB R_{\geq0}$-semimodule, let us call it $\overline{\BB R}_{\geq 0}^X$.
Together with the norm $\norm{\cdot}_\infty$, it can be showed to be a Scott-complete \emph{normed cone} (see~\cite{Selinger2004} or the appendix).
Its cone structure induces an order on it, called the \emph{cone order}:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\overline{\BB R}_{\geq 0}^X$.
Such order actually coincides with the pointwise order on $\overline{\BB R}_{\geq 0}$.
This makes it a Scott-continuous dcpo.
Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
In analogy with \cite[Proposition 17]{DanEhrh2011}, we have:

\begin{theorem}\label{thm:ScottCont}
 All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
 In particular, all tps $f:\overline{\BB R}_{\geq 0}^X\to\overline{\BB R}_{\geq 0}$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
\end{theorem}
%\begin{proof}
% From \autoref{thm:cont} plus the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
%\end{proof}

@@

\begin{proposition}\label{prop:troplinear}
\begin{enumerate}
\item If a tps $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:X\times Y\to \Lawv$ (i.e.~it is tropical linear), then $f$ is non-expansive (i.e.\ $1$-Lipschitz).  
\item If  $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f: \ !_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
\end{enumerate}
\end{proposition}
\begin{proof}[Proof sketch]
1). Using the fact that $f( x)_{b}= \inf_{a\in X}\{\matr f_{a,b}+ x_{a}\}$,
the problem reduces to checking that $|(\matr f_{a,b}- x_{a})- (\matr f_{a,b}- y_{a})| = |x_{a}- y_{a}|\leq \|  x- y\|_{\infty}$ (see the Appendix).
